Declarative Pearl: Lightweight Probabilistic Programming in Curry

### 2015-09-09 Paper/paper.tex
Add comments to conclusion (disadvantage in contrast to HANSEI and in general corresponding to sampling VS exact inference)

% There are many probabilistic languages around these days; a progress
% that started longer than a decade ago, but is currently especially
% active. %
% These languages have their origin either in functional, logic or
% imperative programming, i.e., you can choose a language based on your
% favourite paradigm. %
% Besides newly developed languages, a few libraries for probabilistic
% programming have seen the light of day as well. %
% These languages (stand-alone or DSL) all come with a handful of
% useful tools (functions and stuff) to work with probabilistic values
% and distributions as well as inference algorithms to solve
% probabilistic models that use these values and distributions. %
% That said, there is a large subset that all these languages have in
% common, dispite their different approaches on semantic, syntax as well
% as goals and range of usage, respectively. %
% This subset is rather of practical nature and not of big intereset
% from a PL researcher perspective: every good library or programming
% language needs a handful of primitives and predefined functions, but
% they are not the heart of this piece of software. %

@@

Contributions:
\begin{itemize}
\item We present \emph{PFLP} a lightweight library for probabilistic
  programming in the functional logic language Curry. %
  The library's core consists of ... data structures as well as
  ... functions to work on these strucutres, overall the library needs
  less than 30 lines of code. %
\item We can reduce the library code to such a minimum by reusing Curry's
  semantic. %
  The best part is: no further mechanisms are necessary to enable
  probabilistic computations in Curry. %
  That said, we show high connections between current probability
  programming languages and Curry, including lazy evaluation,
  nondeterministic values as well as corresponding built-in search
  abilities and call-time choice. %
\item We present examples that emphasise the benefits of Curry's
  representation of nondeterminsm, which enables early pruning of the
  search space, compared to Haskell's MonadPlus instance for lists,
  which is often used to model nondeterminism. %
\item It is fair to assume that a wide range of tools existing for
  Curry can be applied to probabilistic languaeges as well. %
  For instance, \cite{Brassel-BIO} has already presented work on a
  debugger for Curry that works well with nondeterminism and free
  variables. %
\end{itemize}

@@

\begin{itemize}
\item abstraction from actual ML problems: define only model and
  questions
\item give simple example for Bayesian Networks (Wet Grass, Alarm,
  etc) and explain functions as superficial as possible
\item point out resemblence between CPT and actual code for model
\end{itemize}
% Experts often state that you do not need to be a rocket scientist in
% order to make sense of research results; however, in the area of
% machine learning, beginners often face difficulties to use the tools
% and current knowledge for their own project. %
% That is, beginners need to invest many hours to digest technical
% details before they can actually make use of top-of-the-art models and
% algorithms. %

% In the field of programming languages, the community often tries to
% generalise and simplify commen concepts and repetative structures. %
% Fortunately, the rising collaboration between ML and PL communities
% has produced a number of programming languages and libraries for
% probabilistic programming that aim at minimising the obstacle to use
% practicle results of ML, thus, empower more users to apply ML
% techniques to their problems. %

@@

After defining the model, we want to ask queries about certian
properties and probabilities, respectively. %
The essence of probabilistic programming is to harmonise a problem and
its analysis: we want to define both, the problem and the queries, in
same language. %

@@

\begin{itemize}
\item PPL vs DSL\\
  Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  % Based on which criteria do you choose the host language? %
  % As for probabilistic languages that started growing over time,
  % libraries that enable probablistic programming exist in the whole
  % variety of paradigms: declarative as well as imperative. %
  % For a start, you obviously choose the paradigm that has the clearest
  % structure and enables straight-forward programming. %
  % We want to address a wide range of developers and even
  % non-PL-experts, thus, we choose a paradigm than can be learnt
  % easily. %
  % Thus, obvioulsy, you choose a declarative programming language. %
  % By choosing Curry, we argue that it is sufficient to reuse built-in
  % features in order to cover the essence of probabilistic programs,
  % especially its semantic. %
\item describe underlying implementation: core data types and
  functions
\item query simple question for example from introduction
\item connect to Martin Erwig's PFP approach
\item Nondeterminism and \code{SearchTree}
\end{itemize}

@@

Rules with overlapping patterns on the left-hand side are one way to
define nondeterministic functions in Curry. %
Alternatively, Curry provides an explicite choice operator \code{?}
that chooses nondeterministicly between its first and second
argument. %
Hence, we can a define a fair coin with the following rule. %

@@

  Such a filter is often called \emph{partial identity} and a common
  functional logic design pattern \cite{funcLogPattern}, additionaly,
  it corresponds to notion of observing a probabilistic value and is a
  standard primitive in probabilistic programming languages. %

\item Each nondeterministic choice corresponds to a branch in the
  search tree. %
  When we evaluate the search tree, we start at the root of the search
  tree and go along a path in that tree: we decide for each branch
  which child to take next. %
  Additionally, we keep track of each decision we make in order to
  repeat a decision in case it occurs several times at different
  positions within the search tree. %
  However, we have to distinct between values that are choosen
  randomly values that share the decision. %
  This distinction is very natural in Curry: a let-binding forces to
  share decisions on probabilistic values. %

@@

  The definition of \code{coin'} consists of an expression with a
  nondeterminstic choice on top-level position, whereas the definition
  of \code{coin} generalises the common parts of the two possible
  values and uses the nondeterminism in a deeper position. %

@@

\section{Conclusion and Future Work}
\begin{itemize}
\item SetFunctions (memory allocation; usage as ``first-class
  citizen'', i.e., check semantics)
\item ``outside'' nondeterminism
\item sampling technique for nondeterministic functions in Curry
\item reevaluate Currys semantic in constrast to Sato's distributuion semantic
\item reactivate debugger for KiCS2/PAKCS
\end{itemize}

>>>>

% There are many probabilistic languages around these days; a progress
% that started longer than a decade ago, but is currently especially
% active. %
% These languages have their origin either in functional, logic or
% imperative programming, i.e., you can choose a language based on your
% favourite paradigm. %
% Besides newly developed languages, a few libraries for probabilistic
% programming have seen the light of day as well. %
% These languages (stand-alone or DSL) all come with a handful of
% useful tools (functions and stuff) to work with probabilistic values
% and distributions as well as inference algorithms to solve
% probabilistic models that use these values and distributions. %
% That said, there is a large subset that all these languages have in
% common, despite their different approaches on semantic, syntax as well
% as goals and range of usage, respectively. %
% This subset is rather of practical nature and not of big interest
% from a PL researcher perspective: every good library or programming
% language needs a handful of primitives and predefined functions, but
% they are not the heart of this piece of software. %

@@

Contributions:
\begin{itemize}
\item We present \emph{PFLP} a lightweight library for probabilistic
  programming in the functional logic language Curry. %
  The library's core consists of ... data structures as well as
  ... functions to work on these structures, overall the library needs
  less than 30 lines of code. %
\item We can reduce the library code to such a minimum by reusing Curry's
  semantic. %
  The best part is: no further mechanisms are necessary to enable
  probabilistic computations in Curry. %
  That said, we show high connections between current probability
  programming languages and Curry, including lazy evaluation,
  nondeterministic values as well as corresponding built-in search
  abilities and call-time choice. %
\item We present examples that emphasise the benefits of Curry's
  representation of nondeterminism, which enables early pruning of the
  search space, compared to Haskell's MonadPlus instance for lists,
  which is often used to model nondeterminism. %
\item It is fair to assume that a wide range of tools existing for
  Curry can be applied to probabilistic languages as well. %
  For instance, \cite{Brassel-BIO} has already presented work on a
  debugger for Curry that works well with nondeterminism and free
  variables. %
\end{itemize}

@@

\begin{itemize}
\item abstraction from actual ML problems: define only model and
  questions
\item give simple example for Bayesian Networks (Wet Grass, Alarm,
  etc) and explain functions as superficial as possible
\item point out resemblance between CPT and actual code for model
\end{itemize}
% Experts often state that you do not need to be a rocket scientist in
% order to make sense of research results; however, in the area of
% machine learning, beginners often face difficulties to use the tools
% and current knowledge for their own project. %
% That is, beginners need to invest many hours to digest technical
% details before they can actually make use of top-of-the-art models and
% algorithms. %

% In the field of programming languages, the community often tries to
% generalise and simplify common concepts and repetitive structures. %
% Fortunately, the rising collaboration between ML and PL communities
% has produced a number of programming languages and libraries for
% probabilistic programming that aim at minimising the obstacle to use
% practical results of ML, thus, empower more users to apply ML
% techniques to their problems. %

@@

After defining the model, we want to ask queries about certain
properties and probabilities, respectively. %
The essence of probabilistic programming is to harmonise a problem and
its analysis: we want to define both, the problem and the queries, in
same language. %

@@

\begin{itemize}
\item PPL vs DSL\\
  Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  % Based on which criteria do you choose the host language? %
  % As for probabilistic languages that started growing over time,
  % libraries that enable probabilistic programming exist in the whole
  % variety of paradigms: declarative as well as imperative. %
  % For a start, you obviously choose the paradigm that has the clearest
  % structure and enables straight-forward programming. %
  % We want to address a wide range of developers and even
  % non-PL-experts, thus, we choose a paradigm than can be learnt
  % easily. %
  % Thus, obviously, you choose a declarative programming language. %
  % By choosing Curry, we argue that it is sufficient to reuse built-in
  % features in order to cover the essence of probabilistic programs,
  % especially its semantic. %
\item describe underlying implementation: core data types and
  functions
\item query simple question for example from introduction
\item connect to Martin Erwig's PFP approach
\item Nondeterminism and \code{SearchTree}
\end{itemize}

@@

Rules with overlapping patterns on the left-hand side are one way to
define nondeterministic functions in Curry. %
Alternatively, Curry provides an explicite choice operator \code{?}
that chooses nondeterministic between its first and second
argument. %
Hence, we can a define a fair coin with the following rule. %

@@

  Such a filter is often called \emph{partial identity} and a common
  functional logic design pattern \cite{funcLogPattern}, additionally,
  it corresponds to notion of observing a probabilistic value and is a
  standard primitive in probabilistic programming languages. %

\item Each nondeterministic choice corresponds to a branch in the
  search tree. %
  When we evaluate the search tree, we start at the root of the search
  tree and go along a path in that tree: we decide for each branch
  which child to take next. %
  Additionally, we keep track of each decision we make in order to
  repeat a decision in case it occurs several times at different
  positions within the search tree. %
  However, we have to distinct between values that are chosen randomly
  values that share the decision. %
  This distinction is very natural in Curry: a let-binding forces to
  share decisions on probabilistic values. %

@@

  The definition of \code{coin'} consists of an expression with a
  nondeterministic choice on top-level position, whereas the
  definition of \code{coin} generalises the common parts of the two
  possible values and uses the nondeterminism in a deeper position. %

@@

\section{Conclusion and Future Work}
\begin{itemize}
\item search tree can only manage weights at leaves, hence, sampling
  is not a feasible alternative to exact inference (that is really
  bad!)
\item Furthermore, a big disadvantage of using Curry in contrast to
  HANSEI is the representation of the search tree. %
  Because of the built-in nondeterminism, we did not have to
  reimplement such a mechanism, but we also have to stick to the
  implementation at is. %
  HANSEI's representation of the search tree has the advantage of
  weighted branches, whereas Curry can only simulate distributions as
  weighted leaves within the search tree. %

\begin{verbatim}
HANSEI
              1.0
        /            \
      0.5            0.5
    /     \        /     \
0.25     0.25    0.3     0.2
\end{verbatim}

\begin{verbatim}
Curry
               .
         /          \
       .             .
    /     \       /      \
0.25     0.25   0.3      0.2
\end{verbatim}

  That is, for exact inference, both approaches have to explore the
  whole search tree, but sampling becomes a lot more easier in HANSEI,
  because at each branch we can prune the search tree by choosing one
  branch and getting rid of all the others. %
  In Curry, we cannot do that, because we do not have any information
  about the probabilities, so in the worst-case we have to explore the
  whole search tree to sample a value. %

\item 
\begin{quote}
  Another point of difference will come out when we talk about memo. %
  We need some sort of state, and that state cannot be global state. %
  When the search is encapsulated, that state has to be be
  encapsulated too, somehow. %
  Hansei also supports so-called nested inference, or nested
  encapsulated search, in Curry terms. %
  That is, when you encapsulate search, some choices are encapsulated
  (represented as nodes in a tree) and some choices are performed. %
  I'm not sure Curry has support for such things. \\%

--- mail correspondence with Oleg Kiselyov
\end{quote}
\end{itemize}

\begin{itemize}
\item SetFunctions (memory allocation; usage as ``first-class
  citizen'', i.e., check semantics)
\item ``outside'' nondeterminism
\item sampling technique for nondeterministic functions in Curry
\item reevaluate Currys semantic in constrast to Sato's distribution
  semantic
\item reactivate debugger for KiCS2/PAKCS
\end{itemize}
